Fusing Filters with Integer Linear Programming

### 2014-04-18 2014betterfusionforfilters/Main.tex
Add notes and move sections around
into: stress filter is just one example of a length changing operation

\begin{abstract}
Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops. However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more. Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle filter-like operators which produce arrays of a different size than the input. We extend the existing ILP approach with support for filters, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-CNF.tex}
\input{section/03-SizeInference.tex}
\input{section/04-ILP.tex}
\input{section/05-Related.tex}
\input{section/XX-Programs.tex}
\section{Conclusion}

>>>>

\begin{abstract}
Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops. However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more. Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different length than their input. We extend the existing ILP approach with support for length changing operators, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-CNF.tex}
\input{section/03-SizeInference.tex}
\input{section/04-ILP.tex}
\input{section/05-Benchmarks.tex}
\input{section/06-Related.tex}
\section{Conclusion}

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
into: stress filter is just one example of a length changing operation

Data flow fusion~\cite{lippmeier2013flow} is a technique to compile a specific class of data flow programs into single, efficient imperative loops. This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}. The key benefits of data flow fusion over prior work are: 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.

@@

If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using two loops: one to determine the sum, and one to divide the elements. The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following example offers \emph{several} unique, but mutually exclusive approaches to fusion. Figure~\ref{f:normalize2-cluterings} on the next page show some of the possibilities.
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us
  = let sum1 = fold   (+)  0   us
        gts  = filter (> 0)    us
        sum2 = fold   (+)  0   gts
        nor1 = map    (/ sum1) us
        nor2 = map    (/ sum2) us
    in (nor1, nor2)
\end{code}

@@

% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
Our contributions are as follows:
\begin{itemize}
\item   
We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for filtering operators. Filters can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).

@@

The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider filtering operations.

>>>>

Data flow fusion~\cite{lippmeier2013flow} is a technique to compile a specific class of data flow programs into single, efficient imperative loops. This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}. The key benefits of data flow fusion over this prior work are: 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.

@@

If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using at least two loops: one to determine the sum, and one to divide the elements. The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following example offers \emph{several} unique, but mutually exclusive approaches to fusion. Figure~\ref{f:normalize2-cluterings} on the next page shows some of the possibilities.
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us
  = let sum1 = fold   (+)  0   us
        gts  = filter (> 0)    us
        sum2 = fold   (+)  0   gts
        nor1 = map    (/ sum1) us
        nor2 = map    (/ sum2) us
    in (nor1, nor2)
\end{code}

@@

% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
Our contributions are as follows:
\begin{itemize}
\item   
We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for length changing operators. Length changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).

@@

The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider length changing operators.
